# Data tansformation

This notebook will cover basics of data transformation using **dply package**

```{r}
library(nycflights13)
library(tidyverse)
```
Let's look at the data
```{r}
flights
```

## dplyr basics

We learn about

- how to pick an observation (**select**)
- how to pick a variable(**filter**)
- create new variable (**mutate**)
- how to summarize many values(**summarize**)
- reorder the row (**arrange**)

Obove functions can be used in conjunction with ***group_by* to allow group by group evaluation
of obove function.

We will aslo see how to apply above  function one after another.

## 5.2 filter rows with filter
Allows subsetting the observation.

To subset jan 1 st flights

```{r}
(jan1 <- filter(flights, month == 1, day == 1 ))
```

### Q 1 
Filter jan 1 flights with carrier is not "UA"

```{r}


```

### Q 2
As you may have gaussed we can use any comparison operator when speficying condition in filter.
But computer are finite precesion machine and we have to be careful when doing floating point arithmetic

Run the foloowing command and explain  the result

```{r}
sqrt(2)^2 == 2
```

Hence use near.

Run following code chunk and see the result.

```{r}
near(sqrt(2)^2, 2)
```

# 5.2.2 Logical operators(&, |, !)

### Q Find all the flight in month 11 or 12

```{r}
filter(flights, )
```

### Q Explain the output of following code

```{r}
filter(flights, month == 11 | 12)
```


To avoid above pitfall use following syntax

```{r}
(nov_dec <- filter(flights, month %in% c(11, 12)))
```


```{r}
(filter(flights, arr_delay <= 120))
```

```{r}
str(flights)
class(flights)
```

# 5.2.3 Missing values or NAs (“not availables”)
**Almost any operaion invloving NA will be NA **.
Use is.na() to check if something is NA. filter command excludes NA.


### Q Find all the values of x greater than 1 or NA in following data frame

```{r}
df = tibble(x= c(1, NA , 3))


```

# 5.2.4 Excercise

#### 1

Find all flights that
- Had an arrival delay of two or more hours
- Were operated by United, American, or Delta
- Departed between midnight and 6am (inclusive)

```{r}
(filter(flights,   ))
```

```{r}
(filter(flights, ))
```

```{r}
(filter(flights,  ))
```


3 How many flights have a missing dep_time? What other variables are missing? What might these rows represent?

```{r}


```


## 5.3 Arrange rows with arrange()

## arrange 
it changes the order of row as per given columns.

```{r}
arrange(flights, year, month, day)
```

**desc()** to re-order by a column in descending order.

```{r}
arrange(flights, desc(arr_delay))
```

## 5.3.1 Exercises

#### Q 1 How could you use arrange() to sort all missing values to the start?

```{r}
df <- tibble(x = c(5, 2, NA))
arrange(df,  )
```
### Q 3 Sort flights to find the fastest flights.

```{r}


```

## 5.4 Select columns with select()

Useful for subsetting the variables in  dataset.

```{r}
select(flights, year, month, day)
```


Or select all the columns between year and day
```{r}
select(flights, year:day)
```

Any gusess what this will do?

```{r}

select(flights, -(year:day))
```

There are a number of helper functions you can use within select():
- starts_with("abc"): matches names that begin with “abc”.
- ends_with("xyz")
- contains("ijk"): matches contain “ijk”.
- matches() selects variables that match a regular expression.
- num_range("x", 1:3) matches x1, x2 and x3.
- everything(): all variables.

e.g

```{r}
select(flights,  ends_with("delay"))
```

But what about everthing?
Helpful if you want to move ...

```{r}
select(flights, )
```


to rename variables.
```{r}
rename(flights, tail_num = tailnum)
```

### Excercise 5.4.1
#### Q 4
Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?


```{r}
select(flights, contains("TIME"))
```

# 5.5 Add new variables with mutate()
How to add new variable? Use mutate. **This helps in feature engineering**

Lets work with smaller datset so that we can see the effect of mutate as it add new variable in the end.
```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)
```


```{r}
mutate(flights_sml,
  gain = arr_delay - dep_delay,
  speed = distance / air_time * 60
)

```
We can refer to columns that you’ve just created.

```{r}
mutate(flights_sml,
  gain = arr_delay - dep_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```


If you only want to keep the new variables, use **transmute()**

```{r}
transmute(flights,
  gain = arr_delay - dep_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```

### 5.5.1 Useful creation functions

- There are many function for creating new  variable with mutate.
Like usual  **arithmetic operator**  and **Modular arithmetic: %/% (integer division) and %% (remainder), log, lo2**  etc.

```{r}
transmute(flights,
  dep_time,
  hour = dep_time %/% 100,
  minute = dep_time %% 100
)
```

- Offsets: lead() and lag()
```{r}
(x<- 1:10)
lead(x)
lag(x)
```
**Q How can above opration be useful**

- Also see  cumsum(), cumprod(), cummin(), cummax(), min_rank, row_rank, dense_rank, percent_rank, cume_rank

## 5.6 Grouped summaries with summarise()

It collapses a data frame to a single row.

```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))

```

summarise() is useful when  we pair it with group_by().

```{r}
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))

```

## 5.6.1 Combining multiple operations with the pipe

Let say we  want to explore the relationship between the distance and average delay for each location.


```{r}
by_dest <- group_by(flights, dest)

delay<- summarise(by_dest, count = n(), dist= mean(distance, na.rm = TRUE),
                  delay = mean(arr_delay, na.rm = TRUE))

delay

delay <- filter(delay, count > 20 , dest != "HNL")

```

Let plot the data to see the trend

```{r}


```

As you can see we have to create a lot of intermediate name, we don't care about.
A much more elegant way is to use **pipe %>%**

```{r}


ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)
```

is someting missing?  ggplot doesn't uses pipe but ggvis uses pipe

## 5.6.2 Missing values

```{r}
flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))
```

Lot of NA.

use na.rm in agregation functions

```{r}
flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay, na.rm = TRUE))

```

or We can first removed cancelled flights

```{r}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))
```

### 5.6.3 Counts

whenever we are doing aggregation, it is always useful to include count, so that we know 
that we are not drawing conclusion based on small amount of data.

Let looks at the planes that have highest average delay.

```{r}
delays <- not_cancelled %>%
  group_by(tailnum) %>%
  summarize(delay = mean(arr_delay))


ggplot(delays, mapping = aes(x = delay))+
  geom_freqpoly(binwidth = 10)

```


Some planes have delay of 300 minutes.
We can drill down more.

```{r}
delays <- not_cancelled %>%
  group_by(tailnum) %>%
  summarise(delay = mean(arr_delay, na.rm = TRUE), n = n())

ggplot(data= delays, mapping = aes(x = n, y = delay))+
  geom_point(alpha = 1/10)
```

Clealy there is much variation in flight delay when there are few flights.

It is always usefull to remove these small sample size so that we can see general tend

```{r}
delays %>%
  filter(n >21) %>%
ggplot(mapping = aes(x =n, y = delay))+
         geom_point(alpha = 1/10)

```


Another variation of this type of problem

```{r}

batting <- as_tibble(Lahman::Batting)
bat_perf <- batting %>%
  group_by(playerID)%>%
  summarise(ba = sum(H, na.rm = TRUE)/sum(AB, na.rm = TRUE),
            ab= sum(AB, na.rm = TRUE))

bat_perf %>%
  filter(ab > 100) %>%
  ggplot(mapping = aes(x= ab, y = ba))+
  geom_point(alpha= 1/10) +
  geom_smooth(se= FALSE)

```
If you naively sort on desc(ba), the people with the best batting averages are clearly lucky, not skilled

```{r}
bat_perf %>%
  arrange(desc(ba))
```



















