---
title: "eda(Exploratory Data Analysis)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

EDA is an iterative cycle

- Ask questions about data
- use visulaization, tranformation and modelling to look for answers
- use insight for new question or refine the questions

Question will helps us in deciding which part of data to focus, grpah to use  etc.

# What question to ask
It will depend on client request and domain.

Two types of questions will always be useful for making discoveries within your data.
- What type of variation occurs within my variables?
- What type of covariation occurs between my variables?

## Visualizing distributions

```{r}
library(ggplot2)

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```
for continous variable what geom functin to use?

```{r}
ggplot(data = diamonds) +
   (mapping = aes(x = carat), binwidth = 0.5)
```

Always play with different binwidth

Here we look on on dimanods with size less than 3 carat

```{r}
library(dplyr)
 diamonds %>% 
  filter(carat < 3) %>%
ggplot(mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.1)

```

In case you want to overlay multiple histogram


```{r}
diamonds %>% 
  filter(carat < 3) %>%
ggplot( mapping = aes(x = carat, colour = cut)) +
  geom_freqpoly(binwidth = 0.1)
```

# Typical and Atypical values
What type of questions we should ask about our plots/data

- common/rare value, why?
- any other pattern

What are the question you can ask about following histogram

```{r}
diamonds %>% 
  filter(carat < 3) %>%
ggplot( mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01)

```
- 
-
-


Clear there is some clustor  based on carat value

What question one can ask about clusters

-
-
-


What kind of information you can infer from following histogram

```{r}
ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  geom_histogram(binwidth = 0.25)
```


# Atypical values or outlier

- they doesn't seem to fit the pattern
- are they error in the process or points to new information

If you have lot of data it can be hard to see outliers


are there some outlier in the following data?


```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)
```



```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50))
```


Let see what kind of obervation they represent

```{r}
unusual <- diamonds %>% 
  filter(y < 3 | y > 20) %>% 
  select(price, x, y, z) %>%
  arrange(y)
unusual
```


One should do analyssis with and wihtout outlier and if they has minimal effect or
we can't figure out why they are there we can replace them, otherwise we shouldn't

drop them wihtout justification. Mention it in your writeup.

#  Missing values
How to handle unsual values
- drop the entire row
```{r}

diamonds2 <- diamonds %>% 
  filter(between(y, 3, 20))
```

What kind of problem we may have in this situation

```{r}

diamonds2 <- diamonds %>% 
  mutate(y = ifelse(y < 3 | y > 20, NA, y))

```

ggplot gives warning

```{r}
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) + 
  geom_point()

```



Some times we want to compare observations with missing values and recorded values.
compare the scheduled departure times for cancelled and non-cancelled times


```{r}
nycflights13::flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(sched_dep_time)) + 
    geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4)


```



# covariation


```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_boxplot()
```











