---
title: "How to use R Markdown"
author: psnegi
date: "01/11/2018"
output:
  html_document: default
---

# Basic
In Rstudio click File - New File - R Markdown to create new R markdown file
Fill the information in header. Its format is like key: value

## Knitting 
It complies the R makrdown file to desired output format. There is small button Knit at the top.
When you click it, it will compile whole Rmakrdown file to selected format.

**How we create R Markdown document**

### R markdown syntax

#### Text formating
One can use \# for level 1 header \#\# for level 2 header etc. to organize the document.

This is plain text. This is *italicized* text. This is **bold** text.
This is  [link](www.du.edu)

#### List
- Unordered list 1
- Unordered list 2
  + subitem 1
  + subitem 2
   
####Table
Table field 1 | Table field 2
--------------| --------------
item 11       | item  12
item 21     | item  13
   
   
#### R code    
*Now more exciting part. How we embed the r or othe language code ?*
We'll use ``` {r } ```
block to embed out code. This is called chunk.

##### An example

````{r}
x <- 1:10
x
````
It display both the code and output

##### Suppressing the code using echo option

```{r echo = FALSE}
x <- 1:10
x
```
##### Suppressing the output using eval option
````{r eval = FALSE}
x <- 1:10
x
````
##### compile the code but include it using include option

````{r eval = FALSE}
x <- 1:10
x
````

#### Using R inline


```{r include = FALSE}

# colorFunc <- "heat.colors"
colorFunc <- "terrain.colors"
# colorFunc <- "topo.colors"
# colorFunc <- "cm.colors"
# colorFunc <- "rainbow"
```

Base R comes with many functions for generating colors. The code below demonstrates the `r colorFunc` function. 


## `r colorFunc`
```{r fig.cap = "The Maunga Whau volcano.", echo = FALSE}
image(volcano, col = get(colorFunc)(200))
```

## "Simple Language Demos"
You can write code in languages other than R with R Markdown, e.g.

## Bash

```{bash}
ls *.Rmd
```



## More "Table options"


Several packages support making beautiful tables with R, such as

* [xtable](https://cran.r-project.org/web/packages/xtable/)
* [stargazer](https://cran.r-project.org/web/packages/stargazer/)
* [pander](http://rapporter.github.io/pander/)
* [tables](https://cran.r-project.org/web/packages/tables/)
* [ascii](http://eusebe.github.io/ascii/)
* etc.

It is also very easy to make tables with knitr's `kable` function:

```{r echo = FALSE, results = 'asis'}
library(knitr)
kable(mtcars[1:5, ], caption = "A knitr kable.")
```

## More "Markdown Demo"

But did you know that you can also use R Markdown's markdown to make

- Latex equations, $E = mc^{2}$
